---
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **eyetools**

## A set of tools for eye data processing, analysis and visualisation in R

**eyetools** is a package that provides a set of simple tools that will facilitate common steps in the processing and analysis of eye data. It is intended for use with data from psychological experiments. The idea is to have a workflow which is aided by these functions, going from processing of the raw data, to extraction of event related data (i.e., fixations, saccades), to summarising those data at the trial level (e.g., time on areas of interest). 

**Warning - still in experimental form! Please check results carefully**

to install: `devtools::install_github("tombeesley/eyetools")`

It is free to use under the GNU General Public Licence..

"Roadmap" for functions:

|order | process | implemented function(s) | comment
|-|-|-|-|
|1.| combine binocular data | `combine_eyes()` | works: either average or "best eye" |
|2.| interpolation | `interpolate()` | working and provides a summary report of repair |
|3.| smoothing | `smoother()` | working |
|4.| dispersion-based fixations  | `fix_dispersion()` | working and pretty fast - needs thorough checking |
|5.| area of interest analysis   | `AOI_time()` | working in basic form - needs thorough checking |
|6.| Visualisations - heatmaps, fixation plots, etc  | `spatial_plot()` | provides a 2D plot of raw data and fixations |
|7.| Saccade detection  | `VTI_saccade()` | Working in basic form - provides summary of velocity, start/end, duration, etc |
|8.| velocity-based fixations | | |
|9.| scan paths?  | | |

## How to use eyetools (work in progress)

```{r include=FALSE}

library(tidyverse)

theme_set(theme_light())


```
**Installation**

You can install eyetools using the following code:
```{r eval=FALSE}
if (!require(devtools)) {
  install.packages("devtools")
  library(devtools)
}
install_github("tombeesley/eyetools")
```
and then load it:
```{r}
library(eyetools)
```

**The format of raw data**

Data needs to be in a particular format to be compatible with the functions in eyetools. This format is 4 columns of data, with each row representing a sample. The four columns are: time, x, y, and trial. You can see an example with the built in data sets:
```{r}
example_raw_sac
```

**Repairing data**

Raw data will often contain missing samples, which we can attempt to repair. eyetools has an `interpolation()` function you can use to do this. It will produce a report of how succesful the repair was in terms of the missing data before and after interpolation:

```{r}
eyetools::interpolate(example_raw_sac, report = TRUE)
raw_data <- eyetools::interpolate(example_raw_sac) # store as new object
```

We can also apply a smoothing function (`smoother()`) over the data, which is particularly important for the analysis of saccadic velocities. 

```{r}
smooth_data <- eyetools::smoother(example_raw_sac) 
```

```{r}
library(tidyverse)

r <- filter(raw_data, trial == 2)
s <- filter(smooth_data, trial == 2)

ggplot() +
  geom_line(data = r, 
            aes(x = time, y = y),
            colour = "red") +
  geom_line(data = s, 
            aes(x = time, y = y),
            colour = "blue")


```

**Processing fixations**

The function `fix_dispersion()` is a dispersion-based algorithm for identifying fixations, based on the algorithm described in Salvucci and Goldberg (2000). Passing raw data to this will return a data frame with the fixations ordered by trial and by fixation sequence, with the averaged x and y coordinates, timestamps and duration. The "min_dur" parameter will restrict to fixations over a certain duration. The "disp_tol" parameter sets the tolerance for the dispersion of data within a fixation. Exploratory analysis of the data will be needed to find suitable values for these.

```{r}

raw_data_f <- filter(raw_data, trial <= 3) # get a sample of trials

fix_dispersion(raw_data_f, min_dur = 120, disp_tol = 100)

```

**Plotting data**

The function `spatial_plot()` is a wrapper for a series of ggplot commands to plot both raw data and fixation summaries. 

```{r fig.height=8, fig.width=8, warning=FALSE}
library(patchwork)
# patchwork is used here to plot adjacent figures

t_raw <- filter(example_raw_sac, trial == 9)

# process fixations
t_fix <- fix_dispersion(t_raw, disp_tol = 100, min_dur = 150)

raw_plot <- spatial_plot(raw_data = t_raw, plot_header = TRUE)
fix_plot <- spatial_plot(raw_data = t_raw, fix_data = t_fix)

raw_plot/fix_plot # combined plot with patchwork

```

**Assessing time on areas of interest**

Example of `AOI_time()` ... 

Show AOI regions on spatial plot...

**Processing saccades**

Example of `VTI_saccade()`

