---
output: 
  md_document:
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "man/figures/")

```

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->


# **eyetools**

## A set of tools for eye data processing, analysis and visualisation in R

**eyetools** is a package that provides a set of simple tools that will facilitate common steps in the processing and analysis of eye data. It is intended for use with data from psychological experiments. The idea is to have a workflow which is aided by these functions, going from processing of the raw data, to extraction of event related data (i.e., fixations, saccades), to summarising those data at the trial level (e.g., time on areas of interest). 

**Warning - still in experimental form! Please check results carefully**

to install: `devtools::install_github("tombeesley/eyetools")`

It is free to use under the GNU General Public Licence..

"Roadmap" for functions:

|order | process | implemented function(s) | comment
|-|-|-|-|
|1.| combine binocular data | `combine_eyes()` | works: either average or "best eye" |
|2.| interpolation | `interpolate()` | working and provides a summary report of repair |
|3.| smoothing | `smoother()` | working |
|4.| dispersion-based fixations  | `fix_dispersion()` | working and pretty fast - needs thorough checking |
|5.| area of interest analysis   | `AOI_time()` | working with both rectangular and circular AOIs |
|6.| Visualisations - heatmaps, fixation plots, etc  | `spatial_plot()` | provides a 2D plot of raw data, fixations and AOIs |
|7.| Saccade detection  | `VTI_saccade()` | Working in basic form - provides summary of velocity, start/end, duration, etc |
|8.| velocity-based fixations | | |
|9.| scan paths ...  | | |

## How to use eyetools (work in progress)

```{r include=FALSE}

library(tidyverse)

theme_set(theme_light())


```
### Installation

You can install eyetools using the following code:
```{r, eval=FALSE}
if (!require(devtools)) {
  install.packages("devtools")
  library(devtools)
}
install_github("tombeesley/eyetools")
```
and then load it:
```{r}
library(eyetools)
```

### The format of raw data

Data needs to be in a particular format to be compatible with the functions in eyetools. This format is 4 columns of data, with each row representing a sample. The four columns are: time, x, y, and trial. You can see an example with the built in data sets:
```{r}
example_raw_sac
```

**Importantly**, the data for each trial should be sequential and logical, that is, the timestamps should run continuously, without any gaps in the recording process. The eyetools package expects data to contain only data for a single participant; none of the functions handle the data for multiple participants in their present state. 

### Repairing data

Raw data will often contain missing samples, which we can attempt to repair. eyetools has an `interpolation()` function you can use to do this. It will produce a report of how successful the repair was in terms of the missing data before and after interpolation:

```{r}
eyetools::interpolate(example_raw_sac, report = TRUE)
raw_data <- eyetools::interpolate(example_raw_sac) # store as new object
```

We can also apply a smoothing function (`smoother()`) over the data, which is particularly important for the analysis of saccadic velocities. 

```{r}
smooth_data <- eyetools::smoother(example_raw_sac) 
```

```{r}
library(tidyverse)

r <- filter(raw_data, trial == 2)
s <- filter(smooth_data, trial == 2)

ggplot() +
  geom_line(data = r, 
            aes(x = time, y = y),
            colour = "red") +
  geom_line(data = s, 
            aes(x = time, y = y),
            colour = "blue")


```

### Processing fixations

The function `fix_dispersion()` is a dispersion-based algorithm for identifying fixations, based on the algorithm described in Salvucci and Goldberg (2000). Passing raw data to this will return a data frame with the fixations ordered by trial and by fixation sequence, with the averaged x and y coordinates, timestamps and duration. The "min_dur" parameter will restrict to fixations over a certain duration. The "disp_tol" parameter sets the tolerance for the dispersion of data within a fixation. Exploratory analysis of the data will be needed to find suitable values for these.

```{r}

raw_data_f <- filter(raw_data, trial <= 3) # get a sample of trials

fix_dispersion(raw_data_f, min_dur = 120, disp_tol = 100)

```



### Plotting data

The function `spatial_plot()` is a wrapper for a series of ggplot commands to plot both raw data and fixation summaries. 

```{r fig.height=8, fig.width=8, warning=FALSE}
library(patchwork)
# patchwork is used here to plot adjacent figures

t_raw <- filter(example_raw_sac, trial == 9)

# process fixations
t_fix <- fix_dispersion(t_raw, disp_tol = 100, min_dur = 150)

raw_plot <- spatial_plot(raw_data = t_raw, plot_header = TRUE)
fix_plot <- spatial_plot(raw_data = t_raw, fix_data = t_fix)

raw_plot/fix_plot # combined plot with patchwork

```

### Assessing time on areas of interest

The function `AOI_time()` can be used to calculate the time spent on areas of interest. Areas of interest need to be defined by the x and y centre points, and the width and height in pixels: 

```{r}

AOI_regions <- data.frame(matrix(nrow = 3, ncol = 4))
colnames(AOI_regions) <- c("x", "y", "width_radius", "height")

AOI_regions[1,] <- c(960, 540, 300, 300) # X, Y, W, H - square
AOI_regions[2,] <- c(200, 540, 300, 300) # X, Y, W, H - square
AOI_regions[3,] <- c(1720, 540, 300, 300) # X, Y, W, H - square

```

`AOI_time()` uses the fixation data as input to the function. In this example we are finding the time spent in 3 rectangular regions across the first 10 trials:

```{r}

t_raw <- filter(example_raw_sac, between(trial,1,10))

# process fixations
t_fix <- fix_dispersion(t_raw, disp_tol = 100, min_dur = 150)

AOI_time(t_fix, AOIs = AOI_regions)

```

We can include the AOIs within our `spatial_plot()`:

```{r}

t_raw <- filter(example_raw_sac, trial == 9) # single trial for plotting purposes

# process fixations
t_fix <- fix_dispersion(t_raw, disp_tol = 100, min_dur = 150)

spatial_plot(raw_data = t_raw, fix_data = t_fix, AOIs = AOI_regions)

```

We can also define AOIs as circles by specifying the radius in the 3rd column and setting the 4th column to NA:

```{r}

AOI_regions <- data.frame(matrix(nrow = 3, ncol = 4))
colnames(AOI_regions) <- c("x", "y", "width_radius", "height")

AOI_regions[1,] <- c(960, 540, 150, NA) # X, Y, R - circle
AOI_regions[2,] <- c(200, 540, 300, 300) # X, Y, W, H - square
AOI_regions[3,] <- c(1720, 540, 300, 300) # X, Y, W, H - square

t_raw <- filter(example_raw_sac, between(trial,1,10))

# process fixations
t_fix <- fix_dispersion(t_raw, disp_tol = 100, min_dur = 150)

spatial_plot(raw_data = t_raw, fix_data = t_fix, AOIs = AOI_regions)

```

Circular AOIs are also handled by AOI_time and will produce different results to comparable rectangular AOIs. Here fixation 5 falls outside of the circular AOI, but within the region of the rectangular AOI:

```{r}

AOI_regions <- data.frame(matrix(nrow = 2, ncol = 4))
colnames(AOI_regions) <- c("x", "y", "width_radius", "height")

AOI_regions[1,] <- c(960, 540, 150, NA) # X, Y, R - circle in centre
AOI_regions[2,] <- c(960, 540, 300, 300) # X, Y, W, H - square in centre

t_raw <- filter(example_raw_sac, trial == 13)

# process fixations
t_fix <- fix_dispersion(t_raw, disp_tol = 100, min_dur = 150)

spatial_plot(raw_data = t_raw, fix_data = t_fix, AOIs = AOI_regions)

AOI_time(t_fix, AOIs = AOI_regions)


```

### Processing saccades

The function `VTI_saccade()` provides a means of processing the data for saccades, based on a "velocity threshold identification" algorithm, as described in Salvucci and Goldberg (2000). As described above, it is wise to use the `smoother()` function on the data first. THe sample rate can be set if known, or can be approximated using the timestamps in the data. The threshold determines the degrees of visual angle per second needed to indicate the presence of a saccadic eye-movement. 

```{r}

t_raw <- filter(example_raw_sac, between(trial,1,10))

t_smooth <- smoother(t_raw)

VTI_saccade(t_smooth, sample_rate = 300)



```

Saccadic eye movements can be plotted alongside other data using the `spatial_plot()` function:

```{r}

t_smooth <- filter(t_smooth, trial == 8)

t_fix <- fix_dispersion(t_smooth, disp_tol = 100, min_dur = 150)

t_sac <- VTI_saccade(t_smooth, sample_rate = 300, threshold = 100)

spatial_plot(raw_data = t_smooth, fix_data = t_fix, sac_data = t_sac)

```




