---
title: "Animating eyetools Data"
author: "Matthew Ivory, Tom Beesley"

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(eyetools)
library(gganimate)
library(tidyverse)
```

In this article, we offer examples of how eyetools data can be used with the [gganimate](https://gganimate.com/) package.

```{r}
data <- combine_eyes(HCL) |> 
  interpolate(participant_ID = "pNum")

data_smooth <- smoother(data, span = .02, participant_ID = "pNum")

```

## Plotting raw data with lag

This animation takes raw data (of a single trial), transforms it into a larger dataset with plot-specific parameters, and then provides an animation of the participant's gaze. The lag feature highlights the path taken and slowly fades to allow more data to be presented

```{r, warning=FALSE}


data_plot1 <- data_smooth |> 
  filter(pNum == "118", trial == "1") |> # take a single participant and trial
  mutate(id = seq(1, n()), # add an ID column to original dataset,
         type = "original",
         size = 10, # specify initial size
         alpha = 1) |> # specify initial alpha
  mutate(time = time - min(time))

#bind the original dataset to a mid and end state too
data_plot1 <- rbind(data_plot1,
                data_plot1 |> # create second version of the dataset - this is the tail
                  mutate(type = "mid_state",      
                         time = time + 45,         
                         size = size * 0.5,        
                         alpha = alpha * 0.2),     
                
                data_plot1 |>                           
                  mutate(type = "end_state",      
                         time = time + 300, # this determines the length of the tail and the final state   
                         size = size * 0.5,        
                         alpha = 0))            

# pass this combined dataset to ggplot & animate
plot_animate <- data_plot1 |>
  ggplot(aes(x = x, y = y, group = id,
             size = size, alpha = alpha)) +
  # add a background image
  annotation_raster(magick::image_read("../data/HCL_sample_image.jpg"),
                    xmin = 0,
                    xmax = 1920,
                    ymin = 0,
                    ymax = 1080) +
  geom_point(colour = "red") +
  transition_components(time) +
  scale_size_identity() +
  scale_alpha_identity() +
  lims(x = c(0, 1920), y = c(0, 1080))

animate(plot_animate,
        duration =  round(max(data_plot1$time)/1000),
        start_pause = 5)

## To save an animation
#anim_save("point_lag.gif", plot_animate, height = 1080, width = 1920, duration = round(max(data_2$time)/1000),
#          end_pause = 5)
```

## Plotting raw and smoothed data

We can also improve upon the static plot presented in the `smoother()` function by animating the x and y together.
```{r}
data_both <- data |> 
  left_join(data_smooth, suffix = c("_raw", "_smooth"), by = join_by("pNum", "time", "trial")) |> 
    filter(pNum == "118", trial == "1") # take a single participant and trial

data_both <- data_both |> 
  pivot_longer(cols = c("x_raw", "x_smooth"), values_to = "x", names_to = "x_names") |> 
  pivot_longer(cols = c("y_raw", "y_smooth"), values_to = "y", names_to = "y_names") |> 
  filter((x_names == "x_raw" & y_names == "y_raw") | (x_names == "x_smooth" & y_names == "y_smooth")) |> 
  mutate(smoothed = str_remove(x_names, "x_"), .after = trial) |> 
  select(-c(x_names, y_names))
```

This first plot is quite rudimentary but shows a single timepoint position and the difference between the raw and `smoother()` data. 

```{r}
plot_animate_smooth <- data_both |> 
  ggplot(aes(x = x, y = y, colour = smoothed)) +
  geom_point(size = 10) +
  lims(x = c(0, 1920), y = c(0, 1080)) +
  transition_components(time)

animate(plot_animate_smooth,
        duration =  round(max(data_both$time)/1000),
        start_pause = 5)

```

### Combining the smooth plot with the lag plot

A much more sophisticated plot is made by combining the lag plot and the `smoother()` comparison plot. In doing so, and tweaking some values, we can see how well the smoothed data adheres to the original data whilst removing the jerkiness of the raw data
```{r}
data_plot3 <- data_both |>
  mutate(id = seq(1, n()), # add an ID column to original dataset,
         type = "original",
         size = 10, # specify initial size
         alpha = 1) |> # specify initial alpha
  mutate(time = time - min(time)) |> 
  arrange(time, desc(smoothed))

#bind the original dataset to a mid and end state too
data_plot3 <- rbind(data_plot3,
                data_plot3 |> # create second version of the dataset - this is the tail
                  mutate(type = "mid_state",      
                         time = time + 45,         
                         size = size * 0.5,        
                         alpha = 1),     
                
                data_plot3 |>                           
                  mutate(type = "end_state",      
                         time = time*2, # this determines the length of the tail and the final state   
                         size = size * 0.5,        
                         alpha = 1))            

data_plot3 <- data_plot3 |> 
  mutate(time = ifelse(time > max(data_both$time), max(data_both$time), time)) |> 
  mutate(alpha = ifelse(type %in% c("mid_state", "end_state") & smoothed == "smooth", .2, alpha))

# pass this combined dataset to ggplot & animate
plot_animate3 <- data_plot3 |>
  ggplot(aes(x = x, y = y, colour = smoothed, group = id,
             size = size, alpha = alpha)) +
  # add a background image
  #annotation_raster(magick::image_read("../data/HCL_sample_image.jpg"),
  #                  xmin = 0,
  #                  xmax = 1920,
  #                  ymin = 0,
  #                  ymax = 1080) +
  geom_point() +
  transition_components(time) +
  scale_size_identity() +
  scale_alpha_identity() +
  lims(x = c(0, 1920), y = c(0, 1080))

animate(plot_animate3,
        duration = round(max(data_plot3$time)/1000),
        start_pause = 5)
```

## Plotting AOI entries over raw data

```{r}
#data <- example_raw_WM |> 
#  interpolate() |> 
#  smoother()
#
#AOIs_WM
#
#data_plot <- data |> 
#  filter(trial == 1) |> 
#  mutate(in_AOI = case_when(between(x, AOIs_WM$x[1], AOIs_WM$x[1] + AOIs_WM$width_radius[1]) ~ TRUE,
#                            between(x, AOIs_WM$x[2], AOIs_WM$x[2] + AOIs_WM$width_radius[1]) ~ TRUE,
#                            between(x, AOIs_WM$x[3], AOIs_WM$x[3] + AOIs_WM$width_radius[1]) ~ TRUE,
#                            between(x, AOIs_WM$x[4], AOIs_WM$x[4] + AOIs_WM$width_radius[1]) ~ TRUE,
#                            .default = FALSE)) |> 
#    mutate(id = seq(1, n()), # add an ID column to original dataset,
#         type = "original",
#         size = 10, # specify initial size
#         alpha = 1) |> # specify initial alpha
#  mutate(time = time - min(time))
#
#
#plot_animate <-  data_plot |>
#  ggplot(aes(x = x, y = y,
#             colour = in_AOI
#             )) +
#  geom_point() +
#  ggtitle("{frame_time}")# + # optional; added to illustrate frame time explicitly
#  transition_components(time)
#
#animate(plot_animate, duration = round(max(data_plot$time)/100))
```
